Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-13 article.tex
transformer-less example of parsers as introduction
simplify parse0 and parse1 code to use standard type classes more; work on refactoring article

\begin{introduction}
Parsing has long been one of the most fascinating and challenging applications
of programming, as evidenced by the large number of parsing libraries available
today.  However, in addition to recognizing a stream of tokens and building a parse
tree, a practical parser has many additional duties.  If parsing fails,
the location and cause of the failure must be accurately reported so that
the user can address the problem
Thus, building composable parsers which accurately report any errors
is extremely difficult.  In this article, we'll see how building
a parsing library around monad transformers allows to build clean and simple 
parsers which nevertheless provide relevant and location-specific error messages.
At the same time, we'll explore the design and some of the tradeoffs of the 
popular Haskell library for monad transformers.
\end{introduction}

\section{Parser Combinators}
Parser combinators have long been used for parsing in the 
functional programming community, represent a powerful and simple solution.
Their main advantages are ability to deal with complicated formats (see
Chomsky levels), language integration, and the declarativeness and
succinctness of the resulting parsers.
Parser combinator libraries typically include a set of basic parsers and a
set of combinators for combining little parsers into big ones.  Parsers are
typically functions whose type is something like:
\begin{verbatim}
[t] -> m (a, [t])
\end{verbatim}
where t is the token type, a is the result type, and m represents some
computational effects.
A very simple parser succeeds, consuming a single token, when the token stream
is not empty, and fails otherwise.  We can make the substitutions t -> Char,
m -> Maybe, and a -> Char and implement the function:
\begin{verbatim}
    item :: [Char] -> Maybe (Char, [Char])
    item (t:ts) = Just (t, ts)
    item []     = Nothing

    -- examples
    ghci$ item "abcde"
    Just ('a', "bcde")

    ghci$ item []
    Nothing
\end{verbatim}

We can also come up with a simple combinator that takes a parser and a function
as input, and succeeds if the parser succeeds and the function applied to the
parser's result succeeds:
\begin{verbatim}
    check :: (a -> Bool) -> ([Char] -> Maybe (a, [Char])) -> [Char] -> Maybe (a, [Char])
    check f p = p >>= \(x, ts) -> if (f x) then Just (x, ts) else Nothing

    -- examples
    ghci$ check (== 'a') item "abcde"
    Just ('a', "bcde")

    ghci$ check (== 'b') item "abcde"
    Nothing
\end{verbatim}
However, in these examples our computational effects are limited to Maybe -- we're
unable to do error reporting, logging, and so on.  Coming up with more general 
parsers and combinators will be the topic of the rest of the article.

\section{Monad transformers}
Monad transformers are a technique for combining monads to form new monads.
The motivation behind them is that while single monads are quite useful, oftentimes
the semantics of multiple monads are required simultaneously; it is then desirable
to have a modular approach for creating combined monads.

There are multiple to approaches to combining monads and to monad transformers
themselves, but we'll follow that of the standard transformers/mtl library, since
it's easy to obtain, easy to use, and quite effective in practice.

@@

@@

Parsing is "the process of analysing a 
string of symbols according to the rules of a formal grammar" (Wikipedia)
The general goal of parsing is to build a tree representing the structure
of the parsed input; further operations such as interpretation or
code generations are then performed on the tree.

@@

\section{Conclusion}
Furthur applications of monad transformers to parsing:  junk logging, partial results, ???

>>>>

\begin{introduction}
Parsing is "the process of analysing a string of symbols according to the rules
of a formal grammar" (Wikipedia), with the general goal of building a tree
representing the structure of the parsed input, upon which further operations
such as interpretation or code generation are then performed.

In addition to recognizing a stream of tokens and building a parse tree, a
real-world parser must recognize faulty input as such, and accurately report the
location and cause of the failure.  But how -- what generates a parse error,
what information is included in the error, how do error-reporting parsers
compose?

Monad transformers provide a clean and simple answer to these questions.
We'll use them to build a parser combinator library, then we'll use the library
build a simple parser, and then we'll add in error reporting.
\end{introduction}

\section{Parser Combinators}
Parser combinators are an excellent approach for building parsers that are easy
to maintain and expressive.  In addition, they benefit from host-language
integration, allowing them to snarf features such as type systems.

Let's make a quick pass through a minimal parser combinator library so that we
can see them in action before we move on to the real stuff.  Without any 
further ado, here's the code:
\begin{verbatim}
import Control.Applicative  (Applicative(..), Alternative(..))
import Control.Monad        (liftM, ap)

newtype Parser t a = 
    Parser {getParser :: [t] -> Maybe ([t], a)}

item :: Parser t t
item = Parser f
  where 
    f [] = Nothing
    f (x:xs) = Just (xs, x)

instance Monad (Parser t) where
  return x = Parser (\ts -> Just (ts, x))
  Parser p >>= f = Parser (\ts ->
                             p ts >>= \(ts', x) ->
                             getParser (f x) ts')

instance Functor (Parser t) where
  fmap = liftM

instance Applicative (Parser t) where
  pure = return
  (<*>) = ap

instance Alternative (Parser t) where
  empty = Parser (const empty)
  Parser l <|> Parser r = Parser (\ts -> l ts <|> r ts)

-- similar to mfilter, but for Alternative instead of MonadPlus
check :: (Monad f, Alternative f) => (a -> Bool) -> f a -> f a
check p m =
    m >>= \x -> if p x then return x else empty

literal :: Eq t => t -> Parser t t
literal x = afilter (== x) item
\end{verbatim}
First, we need to define the Parser type, which says that a Parser is a function
that operates on lists of tokens, producing a new list of tokens and a result
value, and possibly failing.
Next is the item parser, the simplest parser we'll deal with.  All it does is 
consume a single token if available, and succeed otherwise.
For running parsers in sequence, we'll use the combinators from the Monad and
Applicative type classes.  There's a little bit of book-keeping involved in the
Monad instance -- it has to make sure that the correct token list is used by
each parser -- but once we have a Monad instance, the Applicative and Functor
instances can be handled by imports ap and liftM from Control.Monad.
For choosing between alternatives, the MonadPlus type class will do, so we create
an instance that is defined partly in terms of the MonadPlus instance of the
underlying result type, Maybe.
Last are a couple of convenience combinators for creating specific token parsers.

Now for some examples:  first, let's see item in action.  It fails on empty lists,
and succeeds consuming one token otherwise:
\begin{verbatim}
*Main> getParser item ""
Nothing

*Main> getParser item "abcde"
Just ("bcde",'a')
\end{verbatim}
We can run parsers in sequence using the Applicative combinators.  First one parser
is run, then the second.  Note that the token position is handled such that the
second parser sees the token list output by the first parser, and that both parsers
must succeed in order for the entire parser to succeed:
\begin{verbatim}
*Main> getParser (item *> item) "abcde"
Just ("cde",'b')

*Main> getParser (fmap (,) item <*> item) "abcde"
Just ("cde",('a','b'))

*Main> getParser (item *> item) "a"
Nothing
\end{verbatim}
Parsers built out of literal only succeed when the next token matches the given
one:
\begin{verbatim}
*Main> getParser (literal 'a') "abcde"
Just ("bcde",'a')

*Main> getParser (literal 'a') "bcde"
Nothing
\end{verbatim}
Using the MonadPlus combinators, we can create parsers that succeed if any of
their sub-parsers succeed.  This is what allows parsers to backtrack, trying
various alternatives if one fails:
\begin{verbatim}
*Main> getParser (literal 'a' <|> literal 'b') "abcde"
Just ("bcde",'a')

*Main> getParser (literal 'a' <|> literal 'b') "bacde"
Just ("acde",'b')

*Main> getParser (literal 'a' <|> literal 'b') "cdeab"
Nothing
\end{verbatim}
In the rest of this article, we'll be reworking these basic parser combinators.
We'll be simplifying the implementations, extending the types of the results,
all while taking advantage of monad transformers to keep it modular.

@@

\section{Monad transformers}
Monad transformers allow monads to be modularly combined to form new monads
which combine the semantics of their constituents.  That's exactly what we need
to do to build parsers.
We'll try to use the standard transformers/mtl libraries as much as possible, 
since they're easy to obtain, easy to use, and quite effective in practice, but
occasionally we'll have to make an extension to avoid certain type class
instances.

First, note that the parser type we used earlier -- [t] -> Maybe ([t], a) -- 
can be built out of the StateT monad transformer applied to Maybe, so the new
parser type is:
\begin{verbatim}
type Parser t a = StateT [t] Maybe a
\end{verbatim}
And now we can build the item parser using the MonadState combinators:
\begin{verbatim}

\end{verbatim}

@@

@@

\section{Conclusion}
Furthur applications of monad transformers to parsing:  junk logging, partial results, 
 non-deterministic parsing using the ListT or LogicT or just list data type
